---
title: "Decision Trees"
author: "INSOFE Lab Activity on Decision Trees"
date: "23 July 2017"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

# C5.0 Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}
rm(list = ls(all=TRUE))
```

## Goal

* The goal of this activity is to predict wether a patient has liver disease or not based on various patient related attributes


## Agenda 

* Get the data

* Data Pre-processing

* build a model

* predictions

* Communication


## Reading & Understanding the Data

### Read the Data

Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}
getwd()
patient_data<-read.csv("ilpd_data.csv")

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 582 observations and 11 variables

- __The variable descriptions are given below:__

1 - age	 : Age of the patient

2 - gender : Gender of the patient

3 - TB : Total Bilirubin content

4 - DB : Direct Bilirubin content

5 - alk_phos : Alkaline Phosphotase content

6 - alamine : Alamine Aminotransferase content

7 - aspartate : Aspartate Aminotransferase content

8 - TP : Total Protiens content

9 - albumin : 	Albumin content

10 - A/G : Ratio of Albumin and Globulin 

11 - Disease : Whether the patient has liver disease or not 


```{r}
str(patient_data)
summary(patient_data)
```

```{r}
head(patient_data,n=6)
tail(patient_data,n=6)

```


## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}
sum(is.na(patient_data))


```

* Verify the data types assigned to the variables in the dataset

```{r}
str(patient_data)


```

### Split the Data into train and test sets

* Use stratified sampling to split the data into train/test sets (70/30)

* Use the createDataPartition() function from the caret package to do stratified sampling

```{r}
require(lattice)
require(ggplot2)
library(caret)

set.seed(022)
train_row<-createDataPartition(patient_data$disease,p=0.7,list = F)
patient_train<-patient_data[train_row, ]
patient_test<-patient_data[-train_row, ]

```

### Impute the missing values

* Impute missing values using knnImputation() function in both the train and test datasets

```{r}
require(grid)
library(DMwR)

patient_train<-knnImputation(data = patient_train,k=5)
sum(is.na(patient_train))

patient_test<-knnImputation(data=patient_test,k=5)
sum(is.na(patient_test))
```

## build a  Decision Tree

### Model the tree

* Use Quinlan's C5.0 decision tree algorithm implementation from the C50 package to build your decision tree

```{r}
library(C50)

c5_dec_tree<-C5.0(disease ~. , patient_train)
```

* build a rules based tree

```{r}
c5_rule<-C5.0(disease ~., patient_train,rules=T)


```

### Variable Importance in trees

* Find the importance of each variable in the dataset

```{r}

C5imp(c5_dec_tree,metric = "usage")

```

### Rules from trees

* Understand the summary of the returned c5.0 rules based on the decision tree model


```{r}
summary(c5_rule)


```


### Plotting the tree

* Call the plot function on the tree object to visualize the tree

```{r, fig.width= 35, fig.height=15}

plot(c5_dec_tree)

```


## Evaluating the model

### predictions on the test data

* Evaluate the decision tree using the standard error metrics on test data

```{r}

pred_tree<-predict(c5_dec_tree,patient_test)

```

* Report error metrics for classification on test data

```{r}

library(caret)

confusionMatrix(pred_tree,patient_test$disease)


```

# CART Trees

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}
rm(list = ls(all= T))


```

* The classification and regression trees use gini index in place of the gain ratio (based on information gain) used by the ID3 based algorithms, such as c4.5 and c5.0

## Goal

* The goal of this activity is to predict the heating load of a residential building, if the building parameters are given

* Hence, in the future architects would be able to build more energy efficient buildings as they can optimize the building parameters to reduce the heating load

## Agenda 

* Get the data

* Data Pre-processing

* build a model

* predictions

* Communication


## Reading & Understanding the Data

### Read the Data

* Make sure the dataset is located in your current working directory, or else you can change your working directory using the "setwd()" function.

```{r}
getwd()

build<-read.csv("building_energy.csv")

```

### Understand the data

* Use the str(), summary(), head() and tail() functions to get the dimensions and types of attributes in the dataset

* The dataset has 768 observations and 9 variables

```{r}
str(build)
summary(build)

head(build,n=5)
tail(build,n=5)
```

* The variable names are self explanatory, for further information visit http://www.sciencedirect.com/science/article/pii/S037877881200151X

## Data Pre-processing

### Verify Data Integrity

* Verify if the dataset has missing values

```{r}
sum(is.na(build))

```

* Verify the data types assigned to the variables in the dataset

```{r}


str(build)


```

### Split the Data

* Split the data into train/test sets (70/30)

```{r}
library(caret)

set.seed(123)

train_r<-createDataPartition(build$heating_load,p=0.7,list = F)
build_train<-build[train_r, ]
build_test<-build[-train_r, ]
```

## build a Regression Tree

### Model the tree

* Use the rpart package to build a cart tree to predict the heating load

```{r}

library(rpart)

reg_tree<-rpart(heating_load ~., build_train)
printcp(reg_tree)

```

### Tree Explicability

* Print the variable importance

```{r}
reg_tree$variable.importance

```

* Plot the regression tree

```{r, fig.width=8, fig.height=5}

library(rpart.plot)

rpart.plot(reg_tree)


```

## Evaluation on Test Data

* Report error metrics on the test data

```{r}

pred_trees<-predict(reg_tree,build_test)

```

```{r}
library(DMwR)

regr.eval(build_test$heating_load,pred_trees)

```




















